Permutation Pattern matching in\\ $(213,231)$-avoiding permutations

### 2016-10-14 article.tex
a

\begin{abstract}
Given permutations $\sigma$ of size $k$ and $\pi$ of size $n$ with $k<n$, the
\emph{permutation pattern matching} problem is to decide whether $\sigma$ occurs
$\pi$ as an order-isomorphic subsequence.
We give a linear-time algorithm in case both $\pi$ and $\sigma$ avoid
the two size-$3$ permutations $213$ and $231$.
For the special case where only $\sigma$ avoids $213$ and $231$, we present a
$O(max(kn^2,n^2\log \log n)$-time algorithm.
We extend our research to bivincular patterns that avoid $213$ and $231$ and present a $O(kn^4)$-time algorithm.
Finally we look at the related problem of the longest subsequence which avoids $213$ and $231$.
\end{abstract}

@@

A permutation $\sigma$ is said to \emph{occur} in another permutation $\pi$
(or $\pi$ \emph{contains} $\sigma$),
denoted $\sigma \preceq \pi$,
if there exists a subsequence of elements of $\pi$ that has the same relative
order as $\sigma$.
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
For example a permutation contains the permutation $123$ (resp. $321$) if
it has an increasing (resp. a decreasing) subsequence of size $3$.
Similarly, $213$ occurs in $6152347$, but $231$ does not occurs in $6152347$.
During the last decade, the study of patterns in permutations has
become a very active area of research \cite{Kitaev:book:2011} and
a whole annual conference (\textsc{Permutation Patterns}) focuses
on patterns in permutations.

We consider here the so-called \emph{permutation pattern matching} problem
(also sometimes referred to as the \emph{pattern involvement problem}):
Given two permutations $\sigma$ of size $k$ and $\pi$ of size $n$, the problem is to decide whether
$\sigma \preceq \pi$ (the problem is attributed to Wilf in \cite{Bose:Buss:Lubiw:1998}).
The permutation pattern matching is known to be \NPhard~\cite{Bose:Buss:Lubiw:1998}.
It is however polynomial-time solvable by brute-force enumeration
if $\sigma$ has bounded size.
Improvements to this algorithm were presented in
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001} and
\cite{Ahal:Rabinovich:2008},
the latter describing a $O(n^{0.47k+o(k)})$-time algorithm.
Bruner and Lackner \cite{DBLP:journals/corr/abs-1204-5224}
gave a fixed-parameter algorithm solving the permutation pattern matching problem with
an exponential worst-case runtime of $O(1.52^{n}.n.k)$.
This is an improvement upon the $O(k {n \choose k})$ runtime required by
brute-force search without imposing restrictions on $\sigma$ and $\pi$,
where one has to enumerate all the ${n \choose k}$ different subsequences of size $k$ in $\pi$
and test if one of them has the same relative
order as $\sigma$.
Guillemot and Marx \cite{Guillemot:Marx:SODA:2014} showed that the permutation pattern matching problem
is solvable in $2^{O(k^2 \log k)}$, and hence is fixed-parameter tractable with respect to the length $k$
of the pattern $\sigma$ (standard parameterization).
However, Mach proved that the permutation containment problem under the
standard parameterization does not have a polynomial size kernel (assuming $\NPclass \not\subseteq \coNPclass / \POLY$
\cite{Mach:thesis:1992}.

A few particular cases of the permutation pattern matching problem have been attacked successfully.
The case of  increasing patterns is solvable in
$O(n \log \log k)$-time \cite{Crochemore:Porat:2010},
improving the previous 30-year bound of $O(n \log k)$.
% (The algorithm also improves on the previous
% $O(|\pi| \log \log |\pi|)$ bound.)
Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
by stack sorting algorithms.
Any pattern of size $4$ can be detected in $O(n \log n)$ time
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
Algorithmic issues for $321$-avoiding patterns matching for permutations have been investigated in
\cite{Guillemot:Vialette:ISAAC:2009} and more recently in \cite{2015arXiv151006051A}.
The permutation pattern matching problem is also solvable in
polynomial-time for separable patterns \cite{Ibarra:1997,Bose:Buss:Lubiw:1998}
(see also \cite{Bouvel:Rossin:Vialette:CPM:2007} for the longest common subsequence problem and
related ones on separable permutations).
Separable permutations are permutations that do not have an occurrence of
$2413$ nor $3142$, and they are enumerated by the Schr{\"o}der numbers.
%	(sequence A006318 in OEIS).
%	To the best of our knowledge,
%	separable permutations first arose in the work of
%	Avis and Newborn~\cite{Avis:Newborn:1981},
%	who showed that they are precisely the permutations which can be sorted by an
%	arbitrary number of pop-stacks in series,
%	where a pop-stack is a restricted form of stack in which any pop operation
%	pops all items at once.
(Notice that the separable permutations include as a special case the
stack-sortable permutations, which avoid the pattern $231$.)
The major negative result is given in \cite{DBLP:journals/corr/JelinekK16}.
They prove that
the permutation pattern matching problem is \NPC
when $\sigma$ has no decreasing subsequence of length $3$ and
$\pi$ has no decreasing subsequence of length $4$.
This provides the first known example of the permutation pattern matching problem
being hard when one or both of $\pi$ and $\sigma$ are restricted to a proper hereditary
class of permutations.

@@

A \emph{bivincular permutation pattern} $\widetilde{\sigma}$
of size $k$ is a permutation written in
two-line notation
(the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$)
with overlined elements in the top row and underlined elements in the bottom row.
We have the following conditions on the top and bottom rows
of $\sigma$, as seen in \cite{Kitaev:book:2011} in Definition 1.4.1:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the elements corresponding to
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in a occurrence of
	$\sigma$ in $\pi$ must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive elements.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$
	in a permutation $\pi$ must begin with the leftmost
	element of $\pi$,
	and
	if the bottom row of $\sigma$ ends with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$
	in a permutation $\pi$ must end with the rightmost
	element of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the elements corresponding to
	$i, i+1, \ldots, j$ in an
	occurrence of $\sigma$ in $\pi$ must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined
	elements.
	Moreover, if the top row of $\sigma$ begins with
	$^\ulcorner{1}$ then
	any occurrence of $\sigma$ in a permutation $\pi$ must contain
	the bottommost element of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ in a permutation $\pi$ must contain
	the topmost element of $\pi$.
\end{itemize}

@@

As a consequence, every element of a wedge permutation can be partitioned
into three sets: the set of RLMax elements, the set of RLMin elements and the
rightmost element which can be both a RLMax element and a RLMin element.

@@

\begin{remark}
Every element of the set of RLMax elements is above the elements
of the set of the RLMin elements.
In other words, a wedge permutations
can be cut in half horizontally on the rightmost element such that the upper half contains a decreasing
subsequence and the lower half contains an increasing subsequence.
This shape the permutation as a $>$,
hence the name of wedge permutations.
See Figure~\ref{fig:shape of the permutation plus factor}.
\end{remark}

\begin{remark}
The partition gives a bijection between the set of wedge permutations of size $n$ where the elements are $1,\ldots,n$
and the set of binary word of size $n-1$.
The word $w$ which corresponds to $\pi$ is the word where each letter at position $i$
represents whether $\pi[i]$ is a RLMax element or a RLMin element.
We call this bijection $\bijection$.
\end{remark}

To figure out whether or not an element is a RLMax element or a RLMin element,
one has to run the whole permutation from left to right starting
from the position of the element, we give a corollary of lemma~\ref{lemma:first element is 1 or n}, to guest it in constant time.

\begin{corollary}
\label{corollary:max is ascent}
Let $\pi$ be a wedge permutation and $1 \leq i < n$. Then,
\begin{enumerate}
\item $\pi[i]$ is a RLMin element if and only if $\pi[i]<\pi[i+1]$
\item $\pi[i]$ is a RLMax element if and only if $\pi[i]>\pi[i+1]$ is a RLMax
\end{enumerate}

@@

Consider the following problem:
Given a lower bound $\lb$, a upper bound $\ub$, a position i of $\sigma$
and a position $j$ of $\pi$, we want to know if
there exists an occurrence of $\widetilde{\sigma}[i:]$ in $\ptext[j:]$
with every element of the occurrence is in $[\lb,\ub]$
and starting at $\pi[j]$.
In other words given the rectangle with bottom left edge
$(j,\lb)$ and with top right edge $(n,\ub)$
contains an occurrence of $\widetilde{\sigma}[i:]$
starting at $\pi[j]$,
More formally:

$$
\PM{\widetilde{\sigma}}{\ptext}{\lb}{\ub}{i}{j}=
$$
$$
\begin{cases}
	true 	& \text{if $\ptext[j:]$ has an occurrence of the bivincular pattern $\widetilde{\sigma}[i:]$ }\\
			& \text{with every element of the occurrence in $[\lb,\ub]$}\\
			& \text{and starting at $\pi[j]$}\\
			& \text{and if $\overline{\sigma[j](\sigma[j]+1)}$
			 or $\sigma[j]^\urcorner$ appear in $\widetilde{\sigma}$ then $\pi[j]=\ub$} \\
			& \text{and if $\overline{(\sigma[j]-1)\sigma[j]}$
			 or $^\ulcorner{\pi[j]}$ appear in $\widetilde{\sigma}$ then $\sigma[j]=\lb$} \\
	false 	& \text{otherwise}\\
\end{cases}
$$
Clearly if $_\llcorner{\sigma[1]}$ does not appear in $\widetilde{\sigma}$
then $\pi$ contains an occurrence of $\widetilde{\sigma}$
if and only if $\bigcup_{0<j} \PM{\widetilde{\sigma}}{\ptext}{1}{n}{1}{j}$ is true.
If $_\llcorner{\sigma[1]}$ appear in $\sigma$ then
$\pi$ contains an occurrence of
the bivincular wedge permutation pattern $\sigma$
if and only if $\PM{\widetilde{\sigma}}{\ptext}{1}{n}{1}{1}$.

@@

\noindent\textbf{BASE:} \\
$$
\PM{\widetilde{\sigma}}{\ptext}{\lb}{\ub}{k}{j}=
\begin{cases}
	true 	& \text{if $\ptext[j] \in [\lb,\ub ]$}\\
			& \text{and if ${\ppattern[k]}_\lrcorner$ appears in $\widetilde{\sigma}$ then $j=n$}\\
			& \text{and if $_\llcorner{\sigma[k]}$ appears in $\widetilde{\sigma}$ then $j=1$}\\
			& \text{and if ${\ppattern[k]}^\urcorner$ appears in $\widetilde{\sigma}$ then $\ptext[j]=\ub=n$}\\
			& \text{and if  $^\ulcorner{\ppattern[k]}$ appears in $\widetilde{\sigma}$ then $\ptext[j]=\lb=1$ } \\
			& \text{and if  $\overline{(\ppattern[k]-1)\ppattern[k]}$ appears in $\widetilde{\sigma}$ then $\ptext[j]=\lb$ }  \\
			& \text{and if  $\overline{(\ppattern[k]\ppattern[k]+1)}$
			appears in $\widetilde{\sigma}$
			then $\ptext[j]=\ub$}  \\

@@

The base case finds an occurrence for the rightmost element of the pattern. If the rightmost element does not have any restriction on positions and on values, then $\PM{\sigma}{\ptext}{\lb}{\ub}{k}{j}$ is true if and only if $\ppattern[k]$ is matched to $\ptext[j]$. This is true if $\ptext[j] \in [\lb,\ub]$. If ${\ppattern[k]}_\lrcorner$ appears in $\widetilde{\sigma}$ then $\ppattern[k]$ must be matched to the rightmost element of $\pi$ thus $j$ must be $n$. If ${\ppattern[k]}^\urcorner$ appears in $\widetilde{\sigma}$ then $\ppattern[k]$ must be matched to the topmost element which is $n$. If $^\ulcorner{\ppattern[k]}$ appears in $\widetilde{\sigma}$ then $\ppattern[k]$ must be matched to the bottommost element which is $1$.
If  $\overline{(\ppattern[k]-1)\ppattern[k] }$ appears in $\widetilde{\sigma}$ then the matching element of $\ppattern[k]$ and $\ppattern[k]-1$ must be consecutive in value, by recursion the value of the element matching $\ppattern[k]-1$ will be recorded in $\lb$ and by adding $1$ to it thus $\ppattern[k]$ must be matched to $\lb$.
If  $\overline{(\ppattern[k]\ppattern[k]+1)}$ appears in $\widetilde{\sigma}$ then the element matching $\ppattern[k]$ and $\ppattern[k]+1$ must be consecutive in value, by recursion the value of the  element matching $\ppattern[k]+1$ will be recorded in $\ub$ and by removing $1$ to it thus $\ppattern[k]$ must be matched to $\ub$.

@@

In the following, we use the notation $\bigcup$ to
represent the disjunction.
We consider 3 cases for the problem $\PM{ \widetilde{\sigma}}{\ptext}{\lb}{\ub}{i}{j}$:
\begin{itemize}
	\item If $\ptext[j] \notin [\lb,\ub]$ then:
	$$
	\PM{\widetilde{\sigma}}{\ptext}{\lb}{\ub}{i}{j} = false
	$$

@@

	\item If $\ptext[j] \in [\lb,\ub]$ and $\ppattern[i]$ is a RLMin element then:
	$$
	\PM{\widetilde{\sigma}}{\ptext}{\lb}{\ub}{i}{j}=
	\begin{cases}
		\bigcup_{\ell>j} \PM{\widetilde{\sigma}}{\ptext}{\ptext[j]+1}{\ub}{i+1}{\ell}
			& \text{if $\ppattern[i]$ is not underlined } \\
			& \text{ and $\ppattern[i]$ is not overlined} \\
		\bigcup_{\ell>j} \PM{\widetilde{\sigma}}{\ptext}{\ptext[j]+1}{\ub}{i+1}{\ell}
			& \text{if $\ppattern[i]$ is not underlined } \\
			& \text{ and $\overline{(\ppattern[i]-1)\ppattern[i]}$ or $^\ulcorner{\ppattern[i]}$}\\
			& \text{ appears in $\widetilde{\sigma}$}\\
			& \text{ and $\ptext[j]=\lb$} \\
		\PM{\widetilde{\sigma}}{\ptext}{\ptext[j]+1}{\ub}{i+1}{j+1}
			& \text{if $\underline{\ppattern[i]\ppattern[i+1]}$ } \\
			& \text{ appears in $\widetilde{\sigma}$}\\
			& \text{ and $\ppattern[i]$ is not overlined} \\
		\PM{\widetilde{\sigma}}{\ptext}{\ptext[j]+1}{\ub}{i+1}{j+1}
			& \text{if $\underline{\ppattern[i]\ppattern[i+1]}$ } \\
			& \text{ and $\overline{(\ppattern[i]-1)\ppattern[i]}$ or $^\ulcorner{\ppattern[i]}$}\\
			& \text{ appear in $\widetilde{\sigma}$}\\
			& \text{ and $\ptext[j]=\lb$} \\
		false & \text{otherwise}

	\end{cases}
	$$
	Remark that $\ppattern[i]$  can be matched to $\pi[j]$ because $\ptext[j] \in [\lb,\ub]$.
	Thus if $\ptext[j+1:]$ has an occurrence of $\widetilde{\sigma}[i+1:]$  with every element of the occurrence in $[\pi[j]+1,\ub]$ then
 	$\ptext[j:]$ has an occurrence $\widetilde{\sigma}[i:]$.
 	To decide the latter condition,
 	it is enough to know
 	 if there exists $\ell$, $j<\ell$ such that
	$\PM{\widetilde{\sigma}}{\ptext}{\ptext[j]+1}{\ub}{i+1}{\ell}$ is true.
	The first case corresponds to an occurrence without restriction on position and on value.
	The second case asks for the matching of $\ppattern[i]-1$ and $\ppattern[i]$ to be consecutive in value, but the matching of $\ppattern[i]-1$
	is $lb-1$ thus we want $\ptext[j]=\lb$.
	The third case asks for the matching of $\ppattern[i]$ and $\ppattern[i+1]$ to be consecutive in positions, thus the matching of $\ppattern[i+1]$ must be $\pi[j+1]$.
	The fourth case is a union of the second and third case.
%	Note that we do not have to consider the case where $\overline{\ppattern[i](\ppattern[i]+1)}$ appears in $\widetilde{\sigma}$
%	as the element $(\ppattern[i]+1)$ is on the right of $\ppattern[i]$
%	(see Lemma~\ref{lemma:after})
%	and thus will
%	be taken care
%	during the calls of $\PM{\widetilde{\sigma}}{\ptext}{*}{*}{i'}{*}$
%	for some $i<i'$.
%	Identically  we do not have to consider the case
%	where $\sigma[i-1]\sigma[i]$ appears in $\widetilde{\sigma}$

@@

\paragraph{Position Constraint.} There are 3 types of position constraints that can be added by underlined elements.
\begin{itemize}
	\item If $_\llcorner{\sigma[1]}$ appears in $\widetilde{\sigma}$ then the leftmost element of $\sigma$  must be matched to the leftmost element of $\pi$ ($\ppattern[1]$ is matched to $\ptext[1]$ in a occurrence of $\ppattern$ in $\ptext$). This constraint is satisfied by requiring that the occurrence starts at the leftmost element of $\ptext$: if  $\PM{\widetilde{\sigma}}{\ptext}{1}{n}{1}{1}$ is true.

@@

				\item  Every recursive call $\PM{\widetilde{\sigma}}{\ptext}{\lb}{*}{i'}{*}$ is true only if $\sigma[i']$ is matched to the element with value $\lb$.

>>>>

\begin{abstract}
Given permutations $\sigma$ of size $k$ and $\pi$ of size $n$ with $k<n$, the
\emph{permutation pattern matching} problem is to decide whether $\sigma$ occurs
$\pi$ as an order-isomorphic subsequence.
We give a linear-time algorithm in case both $\pi$ and $\sigma$ avoid
the two size-$3$ permutations $213$ and $231$.
For the special case where only $\sigma$ avoids $213$ and $231$, we present a
$O(max(kn^2,n^2\log(\log(n)))$-time algorithm. 
We extend our research to bivincular patterns that avoid $213$ and $231$ and present a $O(kn^4)$-time algorithm.
Finally we look at the related problem of the longest subsequence which avoids $213$ and $231$.  
\end{abstract}

@@

A permutation $\sigma$ is said to occur in another permutation $\pi$,
denoted $\sigma \preceq \pi$,
if there exists a subsequence of elements of $\pi$ that has the same relative
order as $\sigma$.
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
For example the permutation $123$ (resp. $321$) occurs in a second permutation if
the second permutation has an increasing (resp. a decreasing) subsequence of size $3$.
Similarly, $213$ occurs in $6152347$, but $231$ does not occurs in $6152347$.
During the last decade, the study of patterns in permutations has
become a very active area of research \cite{Kitaev:book:2011} and
a whole annual conference (\textsc{Permutation Patterns}) focuses
on patterns in permutations.

We consider here the so-called \emph{permutation pattern matching} problem
(also sometimes referred to as the \emph{pattern involvement problem}):
Given two permutations $\sigma$ of size $k$ and $\pi$ of size $n$, the problem is to decide whether
$\sigma \preceq \pi$ (the problem is attributed to Wilf in \cite{Bose:Buss:Lubiw:1998}).
The permutation pattern matching is known to be \NPhard~\cite{Bose:Buss:Lubiw:1998}.
It is however polynomial-time solvable by brute-force enumeration
if $\sigma$ has bounded size.
Improvements to this algorithm were presented in
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001} and
\cite{Ahal:Rabinovich:2008},
the latter describing a $O(n^{0.47k+o(k)})$-time algorithm.
Bruner and Lackner \cite{DBLP:journals/corr/abs-1204-5224}
gave a fixed-parameter algorithm solving the permutation pattern matching problem with
an exponential worst-case runtime of $O(1.52^{n}.n.k)$.
This is an improvement upon the $O(k {n \choose k})$ runtime required by
brute-force search without imposing restrictions on $\sigma$ and $\pi$,
where one has to enumerate all the ${n \choose k}$ different subsequences of size $k$ in $\pi$
and test if one of them has the same relative
order as $\sigma$.  
A recent major step was taken by Marx and Guillemot
\cite{Guillemot:Marx:SODA:2014}.
They showed that
the permutation pattern matching is fixed-parameter tractable (FPT) for by the size of $\sigma$.

A few particular cases of the permutation pattern matching problem have been attacked successfully.
The case of any increasing pattern is solvable in 
$O(n \log \log k)$ time in the RAM model \cite{Crochemore:Porat:2010},
improving the previous 30-year bound of $O(n \log k)$.
% (The algorithm also improves on the previous
% $O(|\pi| \log \log |\pi|)$ bound.)
Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
by stack sorting algorithms.
Any pattern of size $4$ can be detected in $O(n \log n)$ time
\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
Algorithmic issues for $321$-avoiding patterns matching for permutations have been investigated in
\cite{Guillemot:Vialette:ISAAC:2009} and more recently in \cite{2015arXiv151006051A}.
The permutation pattern matching problem is also solvable in
polynomial-time for separable patterns \cite{Ibarra:1997,Bose:Buss:Lubiw:1998}
(see also \cite{Bouvel:Rossin:Vialette:CPM:2007} for the longest common subsequence problem and related ones on separable permutations).
Separable permutations are permutations that do not have an occurrence of
$2413$ nor $3142$, and they are enumerated by the Schr{\"o}der numbers.
%	(sequence A006318 in OEIS).
%	To the best of our knowledge,
%	separable permutations first arose in the work of
%	Avis and Newborn~\cite{Avis:Newborn:1981},
%	who showed that they are precisely the permutations which can be sorted by an
%	arbitrary number of pop-stacks in series,
%	where a pop-stack is a restricted form of stack in which any pop operation
%	pops all items at once.
(Notice that the separable permutations include as a special case the
stack-sortable permutations, which avoid the pattern $231$.)

@@

A \emph{bivincular permutation pattern} $\widetilde{\sigma}$
of size $k$ is a permutation written in
two-line notation
(the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$)
with overlined elements in the top row and underlined elements in the bottom row.
We have the following conditions on the top and bottom rows
of $\sigma$, as seen in \cite{Kitaev:book:2011} in Definition 1.4.1:
\begin{itemize}
	\item
	If the bottom line of $\widetilde{\sigma}$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the elements corresponding to
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in a occurrence of
	$\sigma$ in $\pi$ must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive elements.
	Moreover if the bottom row of $\widetilde{\sigma}$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\widetilde{\sigma}$
	in a permutation $\pi$ must begin with the leftmost
	element of $\pi$,
	and
	if the bottom row of $\sigma$ ends with
	${\sigma_k}_\lrcorner$ then any occurrence of $\widetilde{\sigma}$
	in a permutation $\pi$ must end with the rightmost
	element of $\pi$.
	\item
	If the top line of $\widetilde{\sigma}$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the elements corresponding to
	$i, i+1, \ldots, j$ in an
	occurrence of $\sigma$ in $\pi$ must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined
	elements.
	Moreover, if the top row of $\widetilde{\sigma}$ begins with
	$^\ulcorner{1}$ then
	any occurrence of $\widetilde{\sigma}$ in a permutation $\pi$ must contain 
	the bottommost element of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\widetilde{\sigma}$ in a permutation $\pi$ must contain
	the topmost element of $\pi$.
\end{itemize}

@@

As a consequence, a wedge permutation can be partitioned 
into three sets: the set of RLMax elements, the set of RLMin elements and the 
rightmost element which can be both a RLMax element and a RLMin element. 
The partition gives a bijection between the set of wedge permutations of size $n$ with elements $1,\ldots,n$
and the set of binary words of size $n-1$.
The word $w$ which corresponds to $\pi$ is the word where each letter at position $i$ 
represents whether $\pi[i]$ is a RLMax element or a RLMin element.
We call this bijection $\bijection$.

@@

\begin{remark}
We can draw a horizontally line (passing thought the rightmost element) on a wedge permutations  such that the upper half contains only a  decreasing  
subsequence and the lower half contains only an increasing subsequence.
This shape the permutation as a $>$, 
hence the name of wedge permutations. 
See Figure~\ref{fig:shape of the permutation plus factor}.
\end{remark}

For any permutation, to figure out whether or not an element is a RLMax element or a RLMin element,
one has to run the whole permutation from left to right starting 
from the position of the element, we give a corollary of lemma~\ref{lemma:first element is 1 or n}, to guest it in constant time for a wedge permutation.

\begin{corollary}
\label{corollary:max is ascent}
Let $\pi$ be a wedge permutation and $1 \leq i < n$. Then,
\begin{enumerate}
\item $\pi[i]$ is a RLMin element if and only if $\pi[i]<\pi[i+1]$
\item $\pi[i]$ is a RLMax element if and only if $\pi[i]>\pi[i+1]$
\end{enumerate}

@@

Consider the following problem:
Given a lower bound $\lb$, a upper bound $\ub$, a position i of $\sigma$
and a position $j$ of $\pi$, we want to know if
there exists an occurrence of $\widetilde{\sigma}[i:]$ in $\ptext[j:]$
with every element of the occurrence is in $[\lb,\ub]$
and starting at $\pi[j]$,
In other words given the rectangle with bottom left edge
$(j,\lb)$ and with top right edge $(n,\ub)$
contains an occurrence of $\widetilde{\sigma}[i:]$
starting at $\pi[j]$,
More formally:

$$
\PM^{\widetilde{\sigma}}_{\ptext}(\lb,\ub,i,j)=
$$
$$
\begin{cases}
	true 	& \text{if $\ptext[j:]$ has an occurrence of the bivincular pattern $\widetilde{\sigma}[i:]$ }\\
			& \text{with every element of the occurrence in $[\lb,\ub]$}\\
			& \text{and starting at $\pi[j]$}\\
			& \text{and if $\overline{\sigma[j](\sigma[j]+1)}$
			 or $\sigma[j]^\urcorner$ appear in $\widetilde{\sigma}$ then $\pi[j]=\ub$} \\
			& \text{and if $\overline{(\sigma[j]-1)\sigma[j]}$
			 or $^\ulcorner{\pi[j]}$ appear in $\widetilde{\sigma}$ then $\sigma[j]=\lb$} \\			 
	false 	& \text{otherwise}\\
\end{cases}
$$
Clearly if $_\llcorner{\sigma[1]}$ does not appear in $\widetilde{\sigma}$
then $\pi$ contains an occurrence of $\widetilde{\sigma}$ 
if and only if $\bigcup_{0<j} \PM^{\widetilde{\sigma}}_{\ptext}(1,n,1,j)$ is true.
If $_\llcorner{\sigma[1]}$ appear in $\sigma$ then 
$\pi$ contains an occurrence of 
the bivincular wedge permutation pattern $\sigma$ 
if and only if $\PM^{\widetilde{\sigma}}_{\ptext}(1,n,1,1)$.

@@

\noindent\textbf{BASE:} \\
$$
\PM^{\widetilde{\sigma}}_{\ptext}(\lb,\ub,k,j)=
\begin{cases}
	true 	& \text{if $\ptext[j] \in [\lb,\ub ]$}\\
			& \text{and if ${\ppattern[k]}_\lrcorner$ appears in $\widetilde{\sigma}$ then $j=n$}\\
			& \text{and if $_\llcorner{\sigma[k]}$ appears in $\widetilde{\sigma}$ then $j=1$}\\			
			& \text{and if ${\ppattern[k]}^\urcorner$ appears in $\widetilde{\sigma}$ then $\ptext[j]=\ub=n$}\\
			& \text{and if  $^\ulcorner{\ppattern[k]}$ appears in $\widetilde{\sigma}$ then $\ptext[j]=\lb=1$ } \\			
			& \text{and if  $\overline{(\ppattern[k]-1)\ppattern[k]}$ appears in $\widetilde{\sigma}$ then $\ptext[j]=\lb$ }  \\
			& \text{and if  $\overline{(\ppattern[k]\ppattern[k]+1)}$ 
			appears in $\widetilde{\sigma}$
			then $\ptext[j]=\ub$}  \\

@@

The base case finds an occurrence for the rightmost element of the pattern. If the rightmost element does not have any restriction on positions and on values, then $\PM^{\sigma}_{\ptext}{\lb}(\ub,k,j)$ is true if and only if $\ppattern[k]$ is matched to $\ptext[j]$. This is true if $\ptext[j] \in [\lb,\ub]$. If ${\ppattern[k]}_\lrcorner$ appears in $\widetilde{\sigma}$ then $\ppattern[k]$ must be matched to the rightmost element of $\pi$ thus $j$ must be $n$. If ${\ppattern[k]}^\urcorner$ appears in $\widetilde{\sigma}$ then $\ppattern[k]$ must be matched to the topmost element which is $n$. If $^\ulcorner{\ppattern[k]}$ appears in $\widetilde{\sigma}$ then $\ppattern[k]$ must be matched to the bottommost element which is $1$. 
If  $\overline{(\ppattern[k]-1)\ppattern[k] }$ appears in $\widetilde{\sigma}$ then the matching element of $\ppattern[k]$ and $\ppattern[k]-1$ must be consecutive in value, by recursion the value of the element matching $\ppattern[k]-1$ will be recorded in $\lb$ and by adding $1$ to it thus $\ppattern[k]$ must be matched to $\lb$. 
If  $\overline{(\ppattern[k]\ppattern[k]+1)}$ appears in $\widetilde{\sigma}$ then the element matching $\ppattern[k]$ and $\ppattern[k]+1$ must be consecutive in value, by recursion the value of the  element matching $\ppattern[k]+1$ will be recorded in $\ub$ and by removing $1$ to it thus $\ppattern[k]$ must be matched to $\ub$.

@@

In the following, we use the notation $\bigcup$ to
represent the disjunction.
We consider 3 cases for the problem $\PM^{\widetilde{\sigma}}_{\ptext}(\lb,\ub,i,j)$:
\begin{itemize}
	\item If $\ptext[j] \notin [\lb,\ub]$ then:
	$$
	\PM^{\widetilde{\sigma}}_{\ptext}(\lb,\ub,i,j) = false
	$$

@@

	\item If $\ptext[j] \in [\lb,\ub]$ and $\ppattern[i]$ is a RLMin element then:
	$$
	\PM^{\widetilde{\sigma}}_{\ptext}(\lb,\ub,i,j)=
	\begin{cases}
		\bigcup_{\ell>j} \PM^{\widetilde{\sigma}}_{\ptext}(\ptext[j]+1,\ub,i+1,\ell)
			& \text{if $\ppattern[i]$ is not underlined } \\
			& \text{ and $\ppattern[i]$ is not overlined} \\
		\bigcup_{\ell>j} \PM^{\widetilde{\sigma}}_{\ptext}(\ptext[j]+1,\ub,i+1,\ell)
			& \text{if $\ppattern[i]$ is not underlined } \\
			& \text{ and $\overline{(\ppattern[i]-1)\ppattern[i]}$ or $^\ulcorner{\ppattern[i]}$}\\
			& \text{ appears in $\widetilde{\sigma}$}\\
			& \text{ and $\ptext[j]=\lb$} \\
		\PM^{\widetilde{\sigma}}_{\ptext}(\ptext[j]+1,\ub,i+1,j+1)
			& \text{if $\underline{\ppattern[i]\ppattern[i+1]}$ } \\
			& \text{ appears in $\widetilde{\sigma}$}\\
			& \text{ and $\ppattern[i]$ is not overlined} \\			
		\PM^{\widetilde{\sigma}}_{\ptext}(\ptext[j]+1,\ub,i+1,j+1)
			& \text{if $\underline{\ppattern[i]\ppattern[i+1]}$ } \\
			& \text{ and $\overline{(\ppattern[i]-1)\ppattern[i]}$ or $^\ulcorner{\ppattern[i]}$}\\
			& \text{ appear in $\widetilde{\sigma}$}\\
			& \text{ and $\ptext[j]=\lb$} \\
		false & \text{otherwise}

	\end{cases}
	$$	
	Remark that $\ppattern[i]$  can be matched to $\pi[j]$ because $\ptext[j] \in [\lb,\ub]$.
	Thus if $\ptext[j+1:]$ has an occurrence of $\widetilde{\sigma}[i+1:]$  with every element of the occurrence in $[\pi[j]+1,\ub]$ then 
 	$\ptext[j:]$ has an occurrence $\widetilde{\sigma}[i:]$. 
 	To decide the latter condition,
 	it is enough to know 
 	 if there exists $\ell$, $j<\ell$ such that
	$\PM^{\widetilde{\sigma}}_{\ptext}(\ptext[j]+1,\ub,i+1,\ell)$ is true.
	The first case corresponds to an occurrence without restriction on position and on value. 
	The second case asks for the matching of $\ppattern[i]-1$ and $\ppattern[i]$ to be consecutive in value, but the matching of $\ppattern[i]-1$
	is $lb-1$ thus we want $\ptext[j]=\lb$. 
	The third case asks for the matching of $\ppattern[i]$ and $\ppattern[i+1]$ to be consecutive in positions, thus the matching of $\ppattern[i+1]$ must be $\pi[j+1]$. 
	The fourth case is a union of the second and third case. 
%	Note that we do not have to consider the case where $\overline{\ppattern[i](\ppattern[i]+1)}$ appears in $\widetilde{\sigma}$
%	as the element $(\ppattern[i]+1)$ is on the right of $\ppattern[i]$
%	(see Lemma~\ref{lemma:after})
%	and thus will
%	be taken care
%	during the calls of $\PM{\widetilde{\sigma}}{\ptext}{*}{*}{i'}{*}$
%	for some $i<i'$.
%	Identically  we do not have to consider the case 
%	where $\sigma[i-1]\sigma[i]$ appears in $\widetilde{\sigma}$

@@

\paragraph{Position Constraint.} There are 3 types of position constraints that can be added by underlined elements.
\begin{itemize} 
	\item If $_\llcorner{\sigma[1]}$ appears in $\widetilde{\sigma}$ then the leftmost element of $\sigma$  must be matched to the leftmost element of $\pi$ ($\ppattern[1]$ is matched to $\ptext[1]$ in a occurrence of $\ppattern$ in $\ptext$). This constraint is satisfied by requiring that the occurrence starts at the leftmost element of $\ptext$: if  $\PM^{\widetilde{\sigma}}_{\ptext}(1,n,1,1)$ is true.

@@

				\item  Every recursive call $\PM^{\widetilde{\sigma}}_{\ptext}(\lb,*,i',*)$ is true only if $\sigma[i']$ is matched to the element with value $\lb$.
